Tracing the Meta-Level: PyPy's Tracing JIT Compiler

### 2009-04-02 talk/icooolps2009/paper.tex
Fix Anto's remarks, add two more references.

A recent approach to getting better performance for dynamic languages is that of
tracing JIT compilers \cite{XXX}. Writing a tracing JIT compiler is relatively
simple, because it can be added to an existing interpreter for a language,
because the interpreter takes over some of the functionality of the compiler and
the machine code generation part can be simplified.
\anto{XXX small english issue: I don't like the two consecutive ``because'',
  I'm not even sure what is the right way to interpret the sentence}

@@

The operations in this sequence are operations of the mentioned intermediate
representation (e.g. note that the generic modulo and equality operations in the
function above have been recognized to always work on integers and are thus
rendered as \texttt{int\_mod} and \texttt{int\_eq}). The trace contains all the
operations that were executed, is in SSA-form \cite{XXX} and ends with a jump
to its own beginning, forming an endless loop that can only be left via a guard
failure. The call to \texttt{f} was inlined into the trace. \sout{Of the condition in
\texttt{f} the much more common \texttt{else} case was traced. The other case is
implemented via a guard failure. This trace can then be turned into machine code
and executed.}
\anto{
Note that the trace contains only the hot \texttt{else} case of the
\texttt{if} test in \texttt{f}, while the other branch is implemented via a
guard failure. This trace can then be turned into machine code and executed.
}

@@

Classical partial evaluation has failed to be useful for dynamic language for
much the same reasons why ahead-of-time compilers cannot compile them to
efficient code. If the partial evaluator knows only the program it simply does
not have enough information to produce good code. Therefore some work has been
done to do partial evaluation at runtime. One of the earliest works on runtime
specialisation is Tempo for C \cite{XXX}. However, it is essentially a normal
partial evaluator ``packaged as a library''; decisions about what can be
specialised and how are pre-determined. Another work in this direction is DyC
\cite{grant_dyc:expressive_2000}, another runtime specializer for C. Both of these projects
have a similar problem as DynamoRIO.  Targeting the C language makes
higher-level specialisation difficult (e.g.\ \texttt{malloc} can not be
optimized).

There has been some attempts to do \emph{dynamic partial evaluation}, which is
partial evaluation that defers partial evaluation completely to runtime
to make partial evaluation more useful for dynamic languages. This concept was
introduced by Sullivan \cite{sullivan_dynamic_2001} who implemented it for a
small dynamic language based on lambda-calculus. There is some work by one of
the authors to implement a dynamic partial evaluator for Prolog \cite{XXX}.

>>>>

A recent approach to getting better performance for dynamic languages is that of
tracing JIT compilers \cite{XXX}. Writing a tracing JIT compiler is relatively
simple. It can be added to an existing interpreter for a language,
the interpreter takes over some of the functionality of the compiler and
the machine code generation part can be simplified.

@@

The operations in this sequence are operations of the mentioned intermediate
representation (e.g. note that the generic modulo and equality operations in the
function above have been recognized to always work on integers and are thus
rendered as \texttt{int\_mod} and \texttt{int\_eq}). The trace contains all the
operations that were executed, is in SSA-form \cite{XXX} and ends with a jump
to its own beginning, forming an endless loop that can only be left via a guard
failure. The call to \texttt{f} was inlined into the trace. Note that the trace
contains only the hot \texttt{else} case of the \texttt{if} test in \texttt{f},
while the other branch is implemented via a guard failure. This trace can then
be turned into machine code and executed.

@@

Classical partial evaluation has failed to be useful for dynamic language for
much the same reasons why ahead-of-time compilers cannot compile them to
efficient code. If the partial evaluator knows only the program it simply does
not have enough information to produce good code. Therefore some work has been
done to do partial evaluation at runtime. One of the earliest works on runtime
specialisation is Tempo for C \cite{consel_general_1996, consel_uniform_1996}.
However, it is essentially a normal
partial evaluator ``packaged as a library''; decisions about what can be
specialised and how are pre-determined. Another work in this direction is DyC
\cite{grant_dyc:expressive_2000}, another runtime specializer for C. Both of these projects
have a similar problem as DynamoRIO.  Targeting the C language makes
higher-level specialisation difficult (e.g.\ \texttt{malloc} can not be
optimized).

There has been some attempts to do \emph{dynamic partial evaluation}, which is
partial evaluation that defers partial evaluation completely to runtime
to make partial evaluation more useful for dynamic languages. This concept was
introduced by Sullivan \cite{sullivan_dynamic_2001} who implemented it for a
small dynamic language based on lambda-calculus. There is some work by one of
the authors to implement a dynamic partial evaluator for Prolog
\cite{carl_friedrich_bolz_automatic_2008}.
